draft

### 2025-05-22 draft.tex
implement shantanu's suggestions

\twocolumn[
\icmltitle{cp\_measure: morphological profiling for data scientists}

@@

\begin{abstract}
Biological image analysis focuses on measuring a particular visual property of interest for cells or other entities. It 
Quantifying object properties in images is a core challenge in biological imaging. The current tools require significant human intervention. Here we introduce an efficient computational library, cp\_measure, which provides programmatic access to the most commonly used metrics to convert images and objects into features. The features are consistent with those produced by the popular graphical user interface-based software CellProfiler, and we showcase tasks for which cp\_measure is more suitable than alternatives. cp\_measure opens the door to community-driven development and improvement of bioimage analysis metrics and pipelines, increasing the scaling capabilities, reproducibility, and accessibility for computational and data scientists.
\end{abstract}
\section{Introduction}
\label{sec:org4c9ba67}
Modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components by microscopy.

Historically, the human eyes and brain drew conclusions from biological images but this approach is limited in scale, objectivity, and reproducibility. Now, quantification is standard, with software often identifying regions of interest (such as cells) and calculating metrics that represent these regions, such as intensity distributions.

Morphological profiling is a technique measuring an array of shape and intensity features for all biological objects, such as cells. These features are fed into statistical or machine learning methods to identify biologically meaningful patterns. One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}. 
\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, for computer-savvy scientists, its graphical user interface and lack of programmatic access means friction in pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.

From the data-science perspective, there are multiple challenges lessening the attractiveness of CellProfiler. First, the integration of CellProfiler with other tools can be a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. Additionally, its usage precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

@@

\section{Extracting interpretable features}
\label{sec:org61842b5}
The library cp\_measure is  the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task. Measurements are categorized based on three kinds of input: One object + one imaging channel (these can be used to measure such metrics as intensity, texture), one object + multiple channels (e.g., for Pearson correlations between intensity values across channels) and multiple objects (e.g., number of neighbours).

By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility. We aim to make the morphological profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent since they are freed from all the orchestration code, we reduced the amount of time and human effort required to integrate it into existing and new pipelines, while still retaining the standard features already extracted for multiple public datasets.

@@

We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images. These correspond to 150 perturbations from the JUMP Cell Painting dataset\citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for a given measurement each. We segmented images to obtain the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler, we calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.

The validation of cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}. We find that most features had a $R^2$ correlation larger than 90\%. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of measurements from our implementation. Some data points fall outside the diagonals for a small set of f, which hint that some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm. This result provides indicates that cp\_measure can be used in some cases instead of CellProfiler, and we would not expect a massive change in results.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocyte 3D nuclei data}
\label{sec:org447090b}

To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow. We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalinin3DCellNuclear2018}. We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells. We achieved 87\% classification accuracy in the testing dataset. Using the morphological profiles, we identified which features distinguish cells based on the time at which the image was taken. Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} Shapley values of the most important features to classify the day of differentiation based on the images (out of three). Each point is an image containing multiple nuclei. The test data accuracy is shown in bold. The x-axis indicates how impactful a given feature is on the classification accuracy. The y-axis are the most significant features, and the last entry is the rest of the features grouped.}
\end{figure}

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. This indicates that the minor axis length of the nuclei is an indicator of phenotypic effect over the course of the experiment, implying that nuclei becomes thicker (though not necessarily more round). Though unsurprising, as we expect astrocyte nuclei to extend over time, by using cp\_measure was able to uncover this with relatively few lines of code.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these measurements as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

@@

Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. We therefore foresee cp\_measure being useful for a broad scientific community beyond morphological profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits: This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the original community for which it was originally developed.

>>>>

\twocolumn[
\icmltitle{cp\_measure: API-first feature extraction for image-based profiling workflows}

@@

\begin{abstract}
Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities. A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles where patterns of changes show biological properties. While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering ML workflows. Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction. We demonstrate that cp\_measure features retain high fidelity with CellProfiler measurements while enabling seamless integration with the scientific Python ecosystem. Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
\end{abstract}

\section{Introduction}
\label{sec:org4c9ba67}
Modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components by microscopy.

Historically, the human eyes and brain drew conclusions from biological images but this approach is limited in scale, objectivity, and reproducibility. Now, quantification is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
We use the terms "features" and "measurements" interchangeably throughout this paper.

Image-based profiling--also termed morphological profiling--is a technique of measuring an array of shape and intensity features for all biological objects, such as cells. These features are fed into statistical or machine learning methods to identify biologically meaningful patterns. One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}.

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, for computer-savvy scientists, its graphical user interface and lack of programmatic access means friction in pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.

From the data-science perspective, there are multiple challenges lessening the attractiveness of CellProfiler. First, coupling CellProfiler with other is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. Its usage also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

@@

\section{Extracting interpretable features}
\label{sec:org61842b5}
The library cp\_measure is  the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task. Measurements are categorized based on three kinds of input: One object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Pearson correlations between intensity values across channels) and multiple objects (e.g., number of neighbours).

By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility. We aim to make the image-based profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent since they are freed from all the orchestration code, we reduced the amount of time and human effort required to integrate it into existing and new pipelines, while still retaining the standard features already extracted for multiple public datasets.

@@

We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images. These correspond to 150 perturbations from the JUMP Cell Painting dataset\citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for a given feature each. We segmented images to obtain the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler, we calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.

The validation of cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}. We find that most features had a $R^2$ correlation larger than 90\%. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of features from our implementation. Some data points fall outside the diagonals for a small set of f, which hint that some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our features have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm. This result provides indicates that cp\_measure can be used in some cases instead of CellProfiler, and we would not expect a massive change in results.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocyte 3D nuclei data}
\label{sec:org447090b}

To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow. We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalinin3DCellNuclear2018}. We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells. We achieved 87\% classification accuracy in the testing dataset. Using the image-based profiles, we identified which features distinguish cells based on the time at which the image was taken. Finally, we calculated the SHAP values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.}

SHAP values of the most important features for classifying the day of differentiation (out of three days). 
Each point represents a single 3D image containing multiple nuclei. 
Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. 
The test data accuracy is shown in bold.}
\end{figure}

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. This indicates that the minor axis length of the nuclei is an indicator of phenotypic effect over the course of the experiment, implying that nuclei becomes thicker (though not necessarily more round). Though unsurprising, as we expect astrocyte nuclei to extend over time, by using cp\_measure was able to uncover this with relatively few lines of code.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these features as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological features yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

@@

Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. We therefore foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits: This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the original community for which it was originally developed.
